Two applications of logic programming to Coq

### 2020-11-07 papers/types2020/elab.tex
done

\item Detailed typing information might not need to be stored within a
  proof since types can often be reconstructed during proof checking.

@@

To give an example, consider using untyped $\lambda$-terms encoded
using De Bruijn's notation as proof certificates for the propositional
intuitionistic logic over just $\supset$.  The fact that such terms
can be used as proof certificates for such formulas (denoting simple
types) can be formally defined using the FPC description in
Figure~\ref{fig:debruijn}.
%
Using the constants provided in that figure, the untyped
$\lambda$-term $\lambda x (x (\lambda y (y (\lambda z (x (\lambda
u\,z))))))$ can be encoded as the following Elpi term of type
\lsti{deb}.
%
\begin{small}
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 [lambda (apply 1 [])])])]))
\end{lstlisting}
\end{small}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.

@@

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
As we have mentioned in Section~\ref{ssec:fpc}, the logic programming
setting allows parallel checking and synthesizing of a pair of
certificates.  That is, during the checking of one certificate, it is
possible to synthesize, for example, a fully detailed term that is
appropriate for handing to the Coq kernel.
If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  (We took exactly this step in Figure~\ref{fig:augmented}
where proof checking involved both proof certificates as well as
Coq-term structures.)  If one is interested in checking only one kind
of external proof structure, then the FPC for that structure could
also be built into the checker (via, say, partial evaluation of logic
programs \cite{lloyd91jlp}).

As an example, consider an external theorem prover for propositional
intuitionistic logic which returns proof structures as untyped
$\lambda$-terms using De Bruijn's notation.   Using the FPC provided
in Figure~\ref{fig:debruijn} and the proof certificate checker in the
file \verb+ljf-dep.mod+ of the repository
\url{https://github.com/proofcert/fpc-elpi/},  the De Bruijn term
display above can be elaborated into a proper proof for the following
Coq theorem.

\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).
\end{lstlisting}

>>>>

\item Detailed typing information might not need to be stored within a
  proof since types can often be reconstructed during proof
  checking~\cite{miller17fac}.

@@

To give an example, consider using untyped $\lambda$-terms encoded
using De Bruijn's notation as proof certificates for propositional
intuitionistic logic over just $\supset$.  The fact that such terms
can be used as proof certificates for such formulas (denoting simple
types) can be formally defined using the FPC description in
Figure~\ref{fig:debruijn}.
%
Using the constants provided in that figure, the untyped
$\lambda$-term $\lambda x (x (\lambda y (y (\lambda z (x (\lambda
u\,z))))))$ can be encoded as the following Elpi term of type
\lsti{deb}.
%
\begin{small}
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 [lambda (apply 1 [])])])]))
\end{lstlisting}
\end{small}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.

@@

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
As we have mentioned in Section~\ref{ssec:fpc}, the logic programming
setting allows parallel checking and synthesizing of a pair of
certificates.  That is, during the checking of one certificate, it is
possible to synthesize, for example, a fully detailed term that is
appropriate for handing to the Coq kernel.
If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  We took exactly this step in Figure~\ref{fig:augmented}
where proof checking involved both proof certificates as well as
Coq terms.  If one is interested in checking only one kind
of external proof structure, then the FPC for that structure could
also be built into the checker (via, say, partial evaluation of logic
programs \cite{lloyd91jlp}).

Continuing with the previous example, consider an external theorem
prover for propositional intuitionistic logic which returns proof
structures as untyped $\lambda$-terms using De Bruijn's notation.
Using the FPC provided in Figure~\ref{fig:debruijn} and the proof
certificate checker in the file \verb+ljf-dep.mod+ of the repository
\url{https://github.com/proofcert/fpc-elpi}, the De Bruijn term
displayed above can be elaborated into a proper proof for the
following Coq theorem.

\begin{lstlisting}[basicstyle=\ttfamily,language=coq]
Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).
\end{lstlisting}
